Impact of Bank of Canada's Monetary Policy on Consumer Debt Levels

### 2024-04-20 paper/paper.qmd
updated the paper

---
title: "Impact of Bank of Canada's Monetary Policy on Consumer Debt Levels"
author: 
  - Abdul Ahad Qureshi
thanks: "Code and data are available at: https://github.com/abdulahadq29/debt."
date: today
date-format: long
abstract: "This analysis investigates the relationship between changes in the Bank of Canada's interest rate policy and consumer debt levels in Canada. Using a dataset of quarterly financial flows and stocks of household credit market debt, the impact of short-term interest rates on credit market debt to disposable income is examined. A linear regression model showed an inverse relationship between short-term interest rates and consumer debt levels. This analysis helps us in understanding the role of fiscal tools in shaping consumer financial behavior."
format: pdf
number-sections: true
bibliograph: references.bib
---

@@

library(ggplot2)
library(dplyr)
library(arrow)
library(here)
library(modelsummary)
#IMPORT ANALYSIS DATA
analysis_data <- read.csv(here("data", "analysis_data", "analysis_data.csv"))

@@

Data analysis for this paper is performed in R [@citeR], and additional help is provided by libraries such as `dplyr` [@dplyr], `ggplot2` [@ggplot], `ggrepel` [@ggrepel], `tidyverse` [@thereferencecanbewhatever], `kableExtra` [@kableextra], `knitr` [@knitr], `haven` [@haven], 'readr' [@readr], `stargazer`[@stargazer], `psych` [@psych], `scales` [@scales], `RColorBrewer` [@rcolorbrewer], `gridExtra` [@gridextra], `here` [@here].

@@

The empirical analysis conducted in this study relies on comprehensive datasets capturing key variables related to consumer debt levels and short-term interest rates in Canada. These datasets provide the empirical foundation for investigating the relationship between changes in the Bank of Canada's interest rate policy and consumer borrowing behavior.

The primary source of data on consumer debt levels is sourced from Statistics Canada. This dataset includes information on various metrics of household indebtedness, including aggregate measures of credit market debt to disposable income, consumer credit, mortgage liabilities, non-mortgage loans, and other relevant indicators. The data cover a substantial time span, allowing for a longitudinal analysis of trends and patterns in consumer borrowing behavior over time.

Complementing the dataset on consumer debt levels is information on short-term interest rates, obtained from FRED. This dataset comprises time-series data on the Bank of Canada's key policy rate, commonly referred to as the overnight rate or the target for the overnight rate. Additionally, the dataset may include data on other short-term interest rates, such as the prime lending rate set by commercial banks, providing a comprehensive view of interest rate movements in the Canadian financial system.

@@

@@

The plot visualizes the trends in the Credit Market Debt to Disposable Income ratio over time, spanning from 1990 to the present day. The x-axis represents the date, while the y-axis depicts the Debt to Disposable Income Ratio, a key metric used to assess the financial health of households.

@@

The plot illustrates the trends in short-term interest rates over time, spanning from 1990 to the present day. The x-axis represents the date, while the y-axis depicts the level of short-term interest rates.

@@

The graph @fig-regression visualizes the correlation between consumer debt and short-term interest rates.

>>>>

---
title: "Impact of Bank of Canada's Monetary Policy on Consumer Debt Levels"
author: 
  - Abdul Ahad Qureshi
thanks: "Code and data are available at: https://github.com/abdulahadq29/debt."
date: today
date-format: long
abstract: "This analysis investigates the relationship between changes in the Bank of Canada's interest rate policy and consumer debt levels in Canada. Using a dataset of quarterly financial flows and stocks of household credit market debt, the impact of short-term interest rates on credit market debt to disposable income is examined. A linear regression model showed an inverse relationship between short-term interest rates and consumer debt levels. This analysis helps us in understanding the role of fiscal tools in shaping consumer financial behavior."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(ggplot2)
library(dplyr)
library(arrow)
library(here)
library(modelsummary)
library(knitr)
#IMPORT ANALYSIS DATA
analysis_data <- read.csv(here("data", "analysis_data", "analysis_data.csv"))

@@

Data analysis for this paper is performed in R [@citeR], and additional help is provided by libraries such as `dplyr` [@dplyr], `ggplot2` [@ggplot], `ggrepel` [@ggrepel], `tidyverse` [@thereferencecanbewhatever], `kableExtra` [@kableextra], `knitr` [@knitr], `haven` [@haven], `readr` [@readr], `RColorBrewer` [@rcolorbrewer], `gridExtra` [@gridextra], `here` [@here], `modelsummary` [@modelsummary]

@@

The empirical analysis conducted in this study relies on comprehensive datasets capturing key variables related to consumer debt levels and short-term interest rates in Canada. These datasets provide the empirical foundation for investigating the relationship between changes in the Bank of Canada's interest rate policy and consumer borrowing behavior. The primary source of data on consumer debt levels is sourced from Statistics Canada. Complementing the dataset on consumer debt levels is information on short-term interest rates, obtained from FRED.

@@

## Data Measurement

In our analysis, two primary metrics form the foundation of the dataset: consumer debt levels and interest rates. Consumer debt is captured as the total debt held by individuals within the domestic economy, encompassing various forms of debt such as credit card balances, student loans, auto loans, and mortgages. The data on consumer debt levels is measured quarterly and reported as an aggregate sum in millions of dollars. This comprehensive approach allows us to observe broader economic patterns and the impact of debt on financial stability.

Interest rates are represented by the central bank’s benchmark short-term rate, which influences borrowing costs throughout the economy. These rates are recorded on a monthly basis and are presented as a percentage, reflecting the cost of borrowing. The interplay between consumer debt levels and interest rates is critical, as fluctuations in rates can significantly affect the ability of consumers to service their debts, thus impacting overall economic health.

In this context, measurement accuracy is paramount. The data utilized in this study have been sourced from reputable financial databases and governmental reports, ensuring reliability and precision in the metrics employed. As economic indicators, both consumer debt and interest rates are subjected to rigorous methods of collection, involving checks and balances to validate their correctness. The data is then standardized to allow for an accurate comparative analysis over time, mitigating inconsistencies that may arise from raw financial figures.

## Key Features

The data on consumer debt information on various metrics of household indebtedness, including aggregate measures of credit market debt to disposable income, consumer credit, mortgage liabilities, non-mortgage loans, and other relevant indicators. The data cover a substantial time span, allowing for a longitudinal analysis of trends and patterns in consumer borrowing behavior over time. The interest rates dataset comprises time-series data on the Bank of Canada's key policy rate, commonly referred to as the overnight rate or the target for the overnight rate. Additionally, the dataset may include data on other short-term interest rates, such as the prime lending rate set by commercial banks, providing a comprehensive view of interest rate movements in the Canadian financial system. A preview of the merged datasets can be seen in @tbl-data.

```{r}
#| label: tbl-data
#| tbl-cap: "A summary table of cleaned data"
#| echo: false
#| eval: true
#| warning: false
#| message: false

data_overview <- head(analysis_data)

# Use kable to create a nice-looking table
kable(data_overview, caption = "Overview of Dataset")

@@

@fig-debt visualizes the trends in the Credit Market Debt to Disposable Income ratio over time, spanning from 1990 to the present day. The x-axis represents the date, while the y-axis depicts the Debt to Disposable Income Ratio, a key metric used to assess the financial health of households.

@@

@fig-interest illustrates the trends in short-term interest rates over time, spanning from 1990 to the present day. The x-axis represents the date, while the y-axis depicts the level of short-term interest rates.

@@

@fig-regression visualizes the correlation between consumer debt and short-term interest rates.
